---
title       : Shiny Calorie Calculator
framework   : revealjs        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow_night      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

## Shiny Calorie Calculator

#### An application that helps you choose a diet plan

Zdravka Cankova

--- 

## Why Count Calories?

One of the most effective ways to lose weight and maintain it
at a healthy level is to follow one simple rule: burn more energy
than you consume. This calorie counter determines how much energy
you burn each day (TDEE), and then helps you set goals for dietary
calories you need to consume to regulate your weight.  

The goals are based on three diet plans:  

1. The first diet plan uses a daily caloric deficit (calories
consumed - calories burned) of 500 calories, resulting in 
approximately 1 lb lost per week.  

2. The second diet plan uses a daily caloric deficit of 1000
calories, resulting in approximately 2 lbs lost per week.  

3. The third diet plan uses a daily caloric deficit calculated
as 20% of TDEE.

The third diet plan is recommended, since its daily calorie 
allowance is easily manageable and is thought to maximize 
fat loss while minimizing muscle loss.

--- 

## Calculating Calories Burned

Below is an example calculation of the TDEE, as well as
dietary calorie consumption recommendations for all 3 diet
plans. The TDEE function has the format:    

TDEE(units, gender, age, height, weight, activity)   

See slide 5 github links for the complete function code. 

```{r, echo = FALSE}
TDEE <- function(units, gender, age, height, weight, activity) {
      if(units == "1"){
            genderless_BMR <- (10 * weight * 0.453592) + (6.25 * height * 2.54) - (5 * age)
      }
      else if(units == "2") {
            genderless_BMR <- (10 * weight) + (6.25 * height) - (5 * age)
      }
      else {stop("Please select units.")}
         
      if(gender == "male") {BMR <- genderless_BMR + 5}
      else if(gender == "female") {BMR <- genderless_BMR - 161}
      else {stop("Please select a gender.")}
      
      if(activity == "sedentary") {TDEE <- round(BMR * 1.2)}
      else if(activity == "light") {TDEE <- round(BMR * 1.375)}
      else if(activity == "moderate") {TDEE <- round(BMR * 1.55)}
      else if(activity == "very active") {TDEE <- round(BMR * 1.725)}
      else if(activity == "extra active") {TDEE <- round(BMR * 1.9)}
      else {stop("Please select activity level.")}
}

TDEE_range <- TDEE("1", "female", 30, 65, 50:500, "light")
```

```{r}
TDEE <- TDEE("1", "female", 30, 65, 160, "light")
# units = "1" selects US units (inches for height and lbs for weight)
TDEE                    # calories burned
TDEE - 500              # first diet calorie allowance
TDEE - 1000             # second diet calorie allowance
round(0.80 * TDEE)      # third diet calorie allowance
```

---

## Comparing Diet Plans

Lastly, the calorie calculator displays an interactive chart that 
allows you to compare the calorie intake for different diet plans 
based on weight. The plot generated for the data in the previous
slide is shown below (see repository linked on slide 5 for
code used):

```{r, echo = FALSE, results = 'asis'}
Lose1lb = TDEE_range - 500
Lose2lb = TDEE_range - 1000
IdealDeficit = round(TDEE_range * 0.80)
cal_vs_weight <- data.frame(Weight = 50:500,
                            "Maintenance" = TDEE_range,
                            "Diet 1" = Lose1lb,
                            "Diet 2" = Lose2lb,
                            "Diet 3" = IdealDeficit,
                            check.names = FALSE)

Chart <- gvisAreaChart(cal_vs_weight,
                      xvar = "Weight",
                      yvar = c("Maintenance",
                               "Diet 1",
                               "Diet 2",
                               "Diet 3"),
                      options = list(height = 400,
                                     width = 900,
                                     chartArea = "{width: '75%'}",
                                     vAxes = "[{title:'Calories',
                                     format:'#,###',
                                     textPosition: 'out'}]",
                                     hAxes = "[{title:'Weight',
                                     textPosition: 'out'}]",
                                     legend = "{position: 'top'}",
                                     areaOpacity = 0.3,
                                     focusTarget = 'category'))

print(Chart, "chart")
```

---

## Links

Link to the Shiny Calorie Calculator:  
https://z-cankova.shinyapps.io/calorie_calculator

Link to Github repository containing source files for the app and this
presentation:  
https://github.com/z-cankova/DevelopingDataProductsPA

Links used to obtain information about calculating TDEE:  
http://www.calculator.net/calorie-calculator.html  
http://www.freedieting.com/calorie_needs.html  
http://www.bmi-calculator.net/bmr-calculator/harris-benedict-equation/  

Links used to obtain diet plan information:  
http://www.acaloriecounter.com/diet/how-many-calories-should-i-eat-per-day-to-lose-weight/  
http://www.acaloriecounter.com/diet/calorie-deficit-to-lose-weight/  
